There once was a grid at ol' Carkeek

### 2016-05-13 Manuscript/v0.0/carkeek_grid.tex
add potential title
misc changes

\title{There once was a grid at ol' Carkeek}

@@

Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples. ?(specifically speed and scope of ecological studies). % "The surveillance of biodiversity using DNA-based detection from environmental samples"
While this approach has been used for decades to survey micro-organisms (CITE VENTER), it has more recently become used as a tool for surveying macro-organisms (CITE EARLY EDNA).
The technique is founded on the premises that:
1) organisms shed DNA-containing material into the environment, 
2) the DNA contained in this material breaks down over time, 
which leads to the conclusion that:
(1) the concentration of DNA originating from an organism should decrease with distance from its source.
%there should be higher concentrations of DNA close to the organism that shed it

@@

For an effectively immobile organism (e.g. fungal hyphae) 
in an effectively static environmental medium (e.g. soil), 
the relationship between DNA concentration and distance appears straightforward (CITE ???).
One might expect the relationship to break down in a dynamic environment: 
An immobile tree may shed cells into the air, which may be carried by the wind for some distance.
This has been a cause for concern for ecologists interested in surveying aquatic organisms, 
because the robustness of eDNA surveys in aquatic environments depends on the distance that DNA travels from a source. 
This distance will be affected by the nature of the environment: 
Material is less likely to be transported (quickly over) great distances in an ice-covered lake 
than along a turbulent shoreline or a fast-moving stream.
One expects more DNA downstream of a fish in a river than upstream, and

In marine environments, this relationship is further complicated by the fact 
that many organisms have dispersive early life history stages (gametes and larvae) 
that can travel great distances but are difficult to track.
These individuals will obscure the expected relationship between DNA concentration and adult abundance because they 
shed genetic material into the environment and 
they are small enough to be captured directly by environmental sampling protocols.
For example, while a traditional survey of a sessile, intertidal bivalve would 
identify the most biomass along the shore, there may be a greater concentration 
of their genetic material at some distance away during a spawning event, or in 
an ocean current carrying many larvae.

% Over two marine habitats, we may KNOW their benthic communities are different, their pelagic/planktonic communities MAY be different, and the DNA community MAY be different.

@@

%-----------------------------------FIGURES-------------------------------------
\section*{Figures}

>>>>

\title{There once was a grid at ol' Carkeek}
% Spatial dispersion of environmental DNA in a marine habitat

@@

Biodiversity surveillance is being revolutionized by DNA-based detection of organisms from environmental samples. ?(specifically speed and scope of ecological studies). % "The surveillance of biodiversity using DNA-based detection from environmental samples"
While this approach has been used for decades to survey micro-organisms (CITE VENTER), it has more recently become used as a tool for surveying macro-organisms (CITE EARLY EDNA).
The technique is founded on the premises that:
(1) organisms shed DNA-containing material into the environment, and 
(2) the DNA contained in this material breaks down over time, 
which leads to the conclusion that the concentration of DNA should decrease with distance from and time since the presence of its source.
%there should be higher concentrations of DNA close to the organism that shed it

@@

For an immobile organism (e.g. fungal hyphae) % effectively 
in a static environmental medium (e.g. soil), %n effectively
the relationship between DNA concentration and distance appears straightforward (CITE ???).
But the relationship may be complicated by a dynamic environment: %One might expect the relationship to break down in 
An immobile tree may shed cells into the air, which may be carried by the wind for some distance.
This has been a cause for concern for ecologists interested in surveying aquatic organisms, 
because the robustness of eDNA surveys for detecting and quantifying organisms depends on the distance that DNA travels from a source. %in aquatic environments 
This distance will be affected by the nature of the environment: 
Material is less likely to be transported (quickly over) great distances in an ice-covered lake 
than along a turbulent shoreline or a fast-moving stream.
One expects more DNA downstream of a fish in a river than upstream, and

% Over two marine habitats, we may KNOW their benthic communities are different, their pelagic/planktonic communities MAY be different, and the DNA community MAY be different.
In marine environments, this relationship is further complicated by the fact 
that many organisms have tiny, highly dispersive early life history stages (gametes and larvae).
%that can travel great distances but are difficult to track.
These individuals will obscure the expected relationship between DNA concentration and adult abundance because they 
shed genetic material into the environment and 
are small enough to be captured directly by environmental sampling protocols.
For example, while a traditional survey of a sessile, intertidal bivalve would 
determine that the most biomass lies along the shore, there may be a greater concentration 
of their genetic material at some distance away during a spawning event, or in 
an ocean current carrying many larvae.

If DNA communities are composed mostly of species known

Despite the potential for discrepancy between communities of benthic organisms and their DNA in the water column above, several studies have reported consistent patterns between the two.

@@

% 1. Jane SF, Wilcox TM, Mckelvey KS, Young MK, Schwartz MK, Lowe WH, et al. Distance, flow and PCR inhibition: EDNA dynamics in two headwater streams. Mol Ecol Resour. 2014; 
%-----------------------------------FIGURES-------------------------------------
\section*{Figures}
